---
title: "Class5. Data Viz with ggplot"
author: "Yixuan Li (PID: A18409882)"
format: pdf
toc: TRUE
---

## Background

There are lots of ways to make plots in R. These include so-called "base R" (like the `plot()` and add on packages like **ggplot**)

Let's make the same plot with these two graphics system. We can use the inbuilt `cars` dataset:

```{r}
head(cars)
```

With "base R" we can simply:

```{r}
plot(cars)
```

Now let's try ggplot, First we need to stall the package using the right code `install.packages("ggplot2")`.

> **N.B.** We never run the `install.packages()` in the code chunck otherwise we need to re-install every time we render our document.

Every time we want to use an add-on package, we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least 3 things:

1.  The **data** i.e. stuff to plot as a data.frame
2.  The **aes** for aesthetics that map the data to the plot
3.  The **geom\_** or geometry i.e. the plot type such as points, lines etc.

```{r}
ggplot(cars) +
  aes(x=speed, y = dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed, y = dist) +
  geom_point() + 
  geom_smooth(method = "gam", formula = y ~ s(x)) +
  labs(x="Speed (MPH)",
       y="Distance (ft)",
       title = "Stopping Distance of Old Cars") +
  theme_light()
```

## Gene Expression Plot

Read some data on the effectes of GLP-1 inhibitor (drug) on gene expression values:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on page.

```{r}
ggplot(genes)+
  aes(Condition1,Condition2)+
  geom_point(col="blue", alpha = 0.2)
```

Let's color by `State` up, down or no change.

```{r}
table(genes$State)
ggplot(genes)+
  aes(x=Condition1,y=Condition2,col = State)+
  geom_point()+
  scale_color_manual(values = c("pink","gray","orange"))+
  labs(x = "Control (no drug)", y="Drug Treatment", title = "Gene Expression Changes Upon Drug Treatment")
```

## Going futher with gapminder

Here we explore the famous `gapminder` dataset with some custom plots.

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Question. How many rows does this dataset have?

```{r}
nrow(gapminder)
```

> Question. How many different continents are in this dataset?

```{r}
table(gapminder$continent)
```

Version 1 plot gdpPercap vs LifeExp for all rows

```{r}
ggplot(gapminder)+
  aes(x = gdpPercap, y = lifeExp, col = continent, size = pop)+
  geom_point(alpha=0.5)
```
Version 2: color the points by value of population:

```{r}
ggplot(gapminder) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.5) +
  scale_size_area(max_size = 10)
```
> Q. produce the plot for 1957 vs 2007

```{r}
library(dplyr)
library(gapminder)

gapminder_1957 <- gapminder %>%
  filter(year %in% c(1957, 2007))

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

##Combining plots

```{r}
library(patchwork)

# Setup some example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      p4
```


















