---
title: "Class18_MiniProject"
author: "Yixuan Li"
format: pdf
editor: visual
---

###**Pertussis and the CMI-PB project**

##Investigating pertussis cases by year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.


```{r}
cdc <- read.csv("cdc.csv")
cdc$No..Reported.Pertussis.Cases <- as.numeric(gsub(",", "", cdc$No..Reported.Pertussis.Cases))

library(ggplot2)

ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(1920, 2020, by = 10)) +
  labs(x = "Year", y = "Number of cases")
```

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, linetype = "dashed") +
  geom_vline(xintercept = 1996, linetype = "dashed") +
  labs(
    x = "Year",
    y = "Number of Pertussis Cases"
  )
```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend? 

After the acellular pertussis (aP) vaccine was introduced in 1996, reported pertussis cases began to increase again after many years of very low incidence. The number of cases gradually rose through the 2000s, with a large spike in 2012, when about 48,277 cases were reported—the highest number since 1955.
Several explanations have been proposed for this resurgence. One possibility is that PCR-based diagnostic tests became more sensitive, allowing more infections to be detected and reported. Another factor may be vaccine hesitancy, which reduces vaccination coverage in some communities. In addition, bacterial evolution may allow the pathogen to partially escape immunity generated by vaccines. Finally, immunity from the aP vaccine may wane more quickly than the older whole-cell (wP) vaccine, especially in adolescents who were vaccinated as infants. These factors together may contribute to the observed increase in pertussis cases


##The CMI-PB API returns JSON data

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```


>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex, subject$race)
```

Working with data

```{r}
library(lubridate)
today()
today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```


>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)

subject$age <- today() - ymd(subject$year_of_birth)

library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round(summary(time_length(ap$age, "years")))

wp <- subject %>% filter(infancy_vac == "wP")

round(summary(time_length(wp$age, "years")))

```


Average age of aP individuals: ~ 26 years

Average age of wP individuals: ~ 36 years

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")

# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

##Joining multiple tables

```{r}
library(jsonlite)
library(dplyr)

# Read specimen table
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)

# Read titer/ab_titer table
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE)
```



>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

# Q9: join specimen and subject
meta <- left_join(specimen, subject, by = "subject_id")

# check dimensions and first few rows
dim(meta)
head(meta)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
library(dplyr)

# Q10: join titer with meta data
abdata <- inner_join(titer, meta, by = "specimen_id")

# check dimensions
dim(abdata)
```


>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

 
```{r}
table(abdata$isotype)
```


>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
# List all unique dataset values
unique(abdata$dataset)

# Count number of rows per dataset
table(abdata$dataset)
```

##Examine IgG Ab titer levels

Now using our joined/merged/linked abdata dataset filter() for IgG isotype.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```



>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
library(ggplot2)

ggplot(igg) +
  aes(x = MFI, y = antigen, fill = biological_sex) +
  geom_boxplot() +
  xlim(0, 75) +
  facet_wrap(vars(visit), nrow = 2)
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?
IgG titers change over time primarily for vaccine-targeted antigens like PT, PRN, and FHA, because these are immunodominant and included in the vaccine, whereas other antigens show little change.



We can attempt to examine differences between wP and aP here by setting color and/or facet values of the plot to include infancy_vac status (see below). However these plots tend to be rather busy and thus hard to interpret easily.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. 

```{r}
library(dplyr)
library(ggplot2)

# Filter for OVA and PT
igg %>%
  filter(antigen %in% c("OVA", "PT")) %>%
  ggplot() +
  aes(x = MFI, y = antigen, col = infancy_vac) +  # MFI on x-axis, antigen on y-axis, color by vaccination type
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
library(dplyr)
library(ggplot2)

igg %>%
  filter(antigen == "FIM2/3") %>%
  ggplot() +
  aes(x = MFI_normalised, y = antigen, col = infancy_vac) +  # normalized MFI on x, antigen on y, color by vaccine type
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```



>Q16. What do you notice about these two antigens time courses and the PT data in particular?

PT shows a vaccine-driven, time-dependent increase in IgG, whereas the control antigen OVA shows no change, highlighting the specificity and effectiveness of the immune response.


>Q17. Do you see any clear difference in aP vs. wP responses?

wP vaccination tends to produce stronger, more sustained IgG responses to PT, whereas aP vaccination leads to lower or more variable responses.

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

>Q18. Does this trend look similar for the 2020 dataset?

The overall trend is very similar between 2020 and 2021 datasets

##Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```


```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

>Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
library(ggplot2)

ggplot(ssrna) +
  aes(x = visit, y = tpm, group = subject_id) +
  geom_point() +
  geom_line(alpha = 0.2) +
  labs(title = "Time course of IGHG1 expression",
       x = "Visit",
       y = "TPM (gene expression)")
```

>Q20. What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

Maximum IGHG1 expression usually occurs earlier than the peak of circulating IgG1 antibody, reflecting that mRNA expression precedes protein production.

>Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

Transcription is upstream of protein production; RNA levels rise rapidly post-stimulation, while circulating antibody levels are delayed due to protein synthesis, secretion, and half-life in plasma.

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```


```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```



