---
title: "Class06: R function"
author: Yixuan Li (A18409882)
format: pdf
toc: true
---

##Background

Functions are at the heart of using R. Everything we do involves calling and using functions (from data input, analysis to results output)

All function in R have at least 3 things:

-   A **name** the thing we use to call the function.
-   One or more input **arguments** that a are comma separated.
-   The **Body**, lines of code between curly brackets { } that does the work of the function.

## A first function

Let's write a wee function that add some numbers:

```{r}
add <- function(x) {
  x + 1
}
```

Let's try it out

```{r}
add(100)
```

Will this work?

```{r}
add(c(100,200,300))
```

Modify to be more useful and add more than just 1

```{r}
add <- function(x , y) {
  x + y
}
```

```{r}
add(100,10)
```


Everything that comma seperated is an argument.
```{r}
plot(1:10, col = "blue", typ="b")
```

If we use log function, the default is natural log. However, we can add an argument to make the base = 10 after a comma.
```{r}
log(10, base = 10)
```

If we change add function and make it function (x , y=1) instead of function (x , y), the default y value would be 1.

```{r}
add <- function(x , y=1) {
  x + y
}
```

> **N.B.** Input arguments can be either **required** or **optional**, The later have a fall-back default that is specified in the function code with euquals sign.

```{r}
#add(100,200,300)
```

### A second function

All functions in R look like this

```
name <- funtion(argument) {
body
}

```

The `sample()` function in R

```{r}
sample(1:100, size = 6)

```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, size = 12, replace = TRUE)
```

> Q. write the code to generate a random 12 nucleotide long DNA sequence.

```{r}
nucleotides <- c("C","G","A","T")
sample(nucleotides,size = 12, replace = TRUE)
```

> Q. Write a first version function called `generate_dna()` that generates a user specified length `n` random DNA sequence?

```{r}
generate_dna <- function(n){
  sample(c("G","T","C","A"),size = n, replace = TRUE)
}
  
```

```{r}
generate_dna(95)
```

> Q. Modify your function to return a FASTA like sequence so rather than  [1] "G" "G" "G" "A" "T". we want "GGGAT". We can ask the LLM to get the result.

```{r}
#Example pattern
x <- c("H","E","L","L","O")
paste(x, collapse = "***")
#returns "HELLO"
paste(x, collapse = "")
```

Here is the code for the answer:
```{r}
generate_dna <- function(n){
  paste(
  sample(c("G","T","C","A"),size = n, replace = TRUE),
  collapse = "")
}
  
```

```{r}
generate_dna(100)
```

Alternative code, name the sample function as "ans":
```{r}
generate_dna <- function(n){
  ans <- sample(c("G","T","C","A"),size = n, replace = TRUE)
  paste(ans, collapse = "")
}
  
```

```{r}
generate_dna(100)
```

> Q. Give the user an option to return FASTA format output sequence or standard multi-element vector format?

Use "flow control", logic control.

```{r}
generate_dna <- function(n=6, fasta = TRUE){
  ans <- sample(c("G","T","C","A"),size = n, replace = TRUE)
  
  if(fasta){
    ans <- paste(ans, collapse = "")
    cat("FASTA format")
  } else{
    cat("Vector format")
  }
  
  return(ans)
}
```

```{r}
generate_dna(10)
generate_dna(10, fasta = FALSE)
```

## A new cool function

> Q. Write a funtion called `generate_protein()` that generates a user specified length protein sequence in FASTA like format?

```{r}
generate_protein <- function(n=6, fasta = TRUE) {
  aa <- c(
  "A","R","N","D","C",
  "E","Q","G","H","I",
  "L","K","M","F","P",
  "S","T","W","Y","V")
   aas <- sample(aa,size = n, replace = TRUE)
  
  if(fasta){
    aas <- paste(aas, collapse = "")
    cat("FASTA format")
  } else{
    cat("Vector format")
  }
  
  return(aas)
  
}
```

```{r}
generate_protein(30)
generate_protein(30, fasta = FALSE)
```

> Q. Use your new `generate_protein()` function to generate all sequences between length 6 and 12 amino-acids in length and check of any of these are unique in nature (i.e. found in the NR database at NCBI)?

```{r}
generate_protein <- function(n=6, fasta = TRUE) {
  aa <- c(
  "A","R","N","D","C",
  "E","Q","G","H","I",
  "L","K","M","F","P",
  "S","T","W","Y","V")
  aas <- sample(aa,size = n, replace = TRUE)
  aas <- paste(aas, collapse = "")
  return(aas)
}
  
```

Create a `for()` loop

```{r}
for(i in 6:12) {
  cat(i, "\n")
  cat(generate_protein(i),"\n")
}
```
FASTA format looks like:

>id
HIAHC

put ">" in code:

```{r}
for(i in 6:12) {
  cat(">",i,sep = "", "\n")
  cat(generate_protein(i),"\n")
}
```
Insert the random sequence in the BLASTp.

The shorter the amino acids sequence is, the less change to find the match. -> unique






























