---
title: "Class13_RNAseq"
author: "Yixuan Li"
format: pdf
editor: visual
toc: TRUE
---

##Set up Bioconductor

Install the package at the console. And then install the DESeq2. Input the library.

```{r}
library(BiocManager)
library(DESeq2)

```

DESeq2 requires two matching data.frames: `countData`, a matrix of raw counts where rows are genes (or other features) and columns are samples (each cell gives the number of reads for gene i in sample j), and `colData`, the sample metadata where rows are samples and columns describe experimental variables (e.g., condition or batch). The sample names in `colData` must exactly match the column names of `countData` to ensure proper alignment for differential expression analysis.

##Import countData and colData

use the `read.csv()` function to read these count data and metadata files.

```{r}
# Complete the missing code
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

Now, take a look at the head of each.

```{r}
head(counts)
```

```{r}
head(metadata)
```

> Q1. How many genes are in this dataset?

```{r}
nrow(counts)
```

> Q2. How many ‘control’ cell lines do we have?

```{r}
sum(metadata$dex == "control")
```

##Toy differential gene expression

Look at the metadata object again to see which samples are `control` and which are drug `treated`.

Note that the control samples are SRR1039508, SRR1039512, SRR1039516, and SRR1039520. This bit of code will first find the sample `id` for those labeled control. Then calculate the mean counts per gene across these samples:

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

> Q3. How would you make the above code in either approach more robust? Is there a function that could help here?

```{r}
control.ids <- metadata$id[metadata$dex == "control"]
control.mean <- rowMeans(counts[ , control.ids])

```

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated.ids <- metadata$id[metadata$dex == "treated"]
treated.mean <- rowMeans(counts[ , treated.ids])

head(treated.ids)
head(treated.mean)
```

We will combine our meancount data for bookkeeping purposes.

```{r}
meancounts <- data.frame(control.mean, treated.mean)

head(meancounts)
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples.

```{r}
plot(meancounts[,1], meancounts[,2],
     xlab = "Control",
     ylab = "Treated")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom\_?() function would you use for this plot?

```{r}
library(ggplot2)

ggplot(meancounts, aes(x = control.mean, y = treated.mean)) +
  geom_point() +
  xlab("Control.mean") +
  ylab("Treated.mean")

```

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?

```{r}
plot(meancounts[,1], meancounts[,2],
     xlab = "Control",
     ylab = "Treated",
     log = "xy")
```

Here we calculate log2foldchange, add it to our `meancounts` data.frame and inspect the results either with the `head()` or the `View()` function for example.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. It turns out that there are a lot of genes with zero expression. Let’s filter our data to remove these genes. Again inspect your result (and the intermediate steps) to see if things make sense to you

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)

```

> Q7. What is the purpose of the `arr.ind` argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

`arr.ind = TRUE` makes `which()` return the row and column positions of TRUE values in a matrix. Here, that tells us which genes (rows) and samples (columns) have zero counts. We take the first column to focus on genes only, and use `unique()` so genes with zeros in multiple samples aren’t counted more than once.

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level?

```{r}
paste("Up:", sum(up.ind))

```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

```{r}
paste("Down:", sum(down.ind))
```

> Q10. Do you trust these results? Why or why not?

No, these results are not fully trustworthy because they are based only on fold change and do not consider statistical significance (e.g., p-values or adjusted p-values). Large fold changes can occur due to variability or low counts, so without proper statistical testing (such as DESeq2), the results may be misleading.

##Setting up for DESeq

```{r}
library(DESeq2)
citation("DESeq2")
```

We use the DESeqDataSetFromMatrix() function to build the required DESeqDataSet object and call it dds, short for our DESeqDataSet. If you get a warning about “some variables in design formula are characters, converting to factors” don’t worry about it. Take a look at the dds object once you create it.

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

##PCA

We will follow the DESeq recommended procedure and associated functions for PCA. First calling `vst()` to apply a variance stabilizing transformation (read more about this in the expandable section below) and then `plotPCA()` to calculate our PCs and plot the results.

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

We can also build the PCA plot from scratch using the ggplot2 package. This is done by asking the plotPCA function to return the data used for plotting rather than building the plot.

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

Calculate percent variance per PC for the plot axis labels

```{r}
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

##DESeq analysis

`dds` is a bare-bones DESeqDataSet. The `DESeq()` function takes a DESeqDataSet and returns a DESeqDataSet, but with additional information filled in. Notice how if we try to access these results before running the analysis, nothing exists.

we’re running the DESeq pipeline on the `dds` object, and reassigning the whole thing back to `dds`, which will now be a DESeqDataSet populated with all those values.

```{r}
dds <- DESeq(dds)
```

Getting results

```{r}
res <- results(dds)
res
```

We can summarize some basic tallies using the summary function.

```{r}
summary(res)
```

The results function contains a number of arguments to customize the results table. By default the argument alpha is set to 0.1. If the adjusted p value cutoff will be a value other than 0.1, alpha should be set to that value:

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

##Adding annotation data

We will use one of Bioconductor’s main annotation packages to help with mapping between various ID schemes. Here we load the AnnotationDbi package and the annotation data package for humans org.Hs.eg.db.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

The later of these is is the organism annotation package (“org”) for Homo sapiens (“Hs”), organized as an AnnotationDbi database package (“db”), using Entrez Gene IDs (“eg”) as primary key. To get a list of all available key types that we can use to map between, use the columns() function:

```{r}
columns(org.Hs.eg.db)

```

The main function we will use from the AnnotationDbi package is `called mapIds()`. We can use the `mapIds()` function to add individual columns to our results table.

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res\$genename.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     column = "ENTREZID",
                     keytype = "ENSEMBL",
                     multiVals = "first")

res$uniprot <- mapIds(org.Hs.eg.db,
                      keys = row.names(res),
                      column = "UNIPROT",
                      keytype = "ENSEMBL",
                      multiVals = "first")

res$genename <- mapIds(org.Hs.eg.db,
                       keys = row.names(res),
                       column = "GENENAME",
                       keytype = "ENSEMBL",
                       multiVals = "first")

head(res)
```

You can arrange and view the results by the adjusted p-value

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

Write out the results

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

##Data Visualization

Volcano plots

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

To make this more useful we can add some guidelines (with the abline() function) and color (with a custom color vector) highlighting genes that have padj\<0.05 and the absolute log2FoldChange\>2.

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="pink", lty=2)
abline(h=-log(0.05), col="pink", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "orange" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "pink"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

For even more customization you might find the EnhancedVolcano bioconductor package useful (Note. It uses ggplot under the hood):

First we will add the more understandable gene symbol names to our full results object res as we will use this to label the most interesting genes in our final plot.

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```
