---
title: "Class07: Machine Learning 1"
author: "Yixuan Li"
format: pdf
toc: true
---

##Background Today we will begin our exploration of some important machine learing methods, namely **clustering** and **dimensionality reduction**.

Let's make up some input data for clustering where we know what the nature "clusters" are.

The function `rnorm()` can be useful here

```{r}
hist(rnorm(5000))
```

> Q. Generate 30 random numbers centered at +3 and another 3- centered at -3.

```{r}
tmp <- c(rnorm(30, mean=3),
         rnorm(30, mean=-3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

## K-means clustering

The main function in "base R" for K-means clustering is called `kmeans()`:

```{r}
km <- kmeans(x, centers=2)
km
```

> Q. What component of the results object details the cluster sizes

```{r}
km$size
```

> Q. What component of the results object details the cluster center

```{r}
km$center
```

> Q. What component of the results object the cluster membership vector (i.e. our main result of which ppints lie in which cluster)

```{r}
km$cluster
```

> Q. Plot our clustering results with points colored by cluster and also the cluster centers as new points colored blue

```{r}
plot(x, col=c(1,2))
points(km$centers, col="blue", pch=15)
```

> Q. Run `kmeans()` again and this time produce 4 clusters (and call your result object `k4`) and make a results figure like above?

```{r}
k4 <- kmeans(x, centers=4)
k4
```

```{r}
plot(x, col=c(1,2,3,4))
points(k4$centers, col="blue", pch=15)
```

The metric

```{r}
#total within-cluster sum of squares. It measure how tightly the data points are grouped inside their clusters.
km$tot.withinss
k4$tot.withinss
```

> Q. Let's try different number of K (centers) from 1 to 30 and see what the best result is?

```{r}
#create an loop. start with an NULL for default ans value
ans <- NULL
for(i in 1:30){
ans <- c(ans, kmeans(x, centers = i)$tot.withinss)
#Store the value in the ans
}
#plot the answers
plot(ans, typ = "o")
```

The best cluster number is 2. 

**Key-point** K-means will impose a clustering structure on your data even if it is not there - it will always give you the answer you asked for even if that answer is silly.

## Hierarchiral Clustering

The main function for Hierarchical Clustering is called `hclust()`
Unlike `kmeans()` (which does all the work for you) you can't just pass `hclust()` our raw input data.It needs a "distance matrix
like rhe one returned from the `dist()` function

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

To extract our cluster membership vector from a `hclust()` result object we have to "cut" our tree at a given height to yield seperate "groups"/"branches"

```{r}
plot(hc)
abline(h=8, col="red",lty=2)
```

To do this we use the `cutree()` function on our `hclust()` object:

```{r}
grps <- cutree(hc, h=8)
```

```{r}
table(grps, km$cluster)
```
## PCA for UK food data

Import the dataset of food consumption in the UK:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns?

```{r}
dim(x)
```

One solution to set the row names is to do it by hand.
```{r}
#make the first column an rownames and get rid of the data.
#rownames(x)
rownames(x) <- x[ ,1]
```

To remove the first column I can use the minus index trick.

```{r}
x <- x[ ,-1]
# This is not good because we lose the first column every time we run it and eventually all the column will be lost.
```

Better way to do this is to set the row names to the first column with `read.csv()`

```{r}
x <- read.csv(url, row.names = 1)
x
```
```{r}
plot(x)
```

### Spotting major differences and trends

Difficult to show the plots 

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

Heat map

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )

```

For heat map, it's confusing what comparison it did. which 2 data it compares.

##PCA to the rescue

The main PCA function in "base R" is called `prcomp()`, This function wants the transpose of our food data as input (i. e. the foods as columns and the countries as rows)

```{r}
pca <- prcomp (t(x))
# Try to find the best component that will do the best job to deal with the data
```


```{r}
summary(pca)
```
```{r}
attributes(pca)
```
To make one of main PCA result figures we turn to `pca$x()` the scores along our new PCs. This is called "PC plot" or "score plot" or "Ordienation plot"...
```{r}
pca$x
```

```{r}
my_cols <- c("orange","red","blue","darkgreen")
```

```{r}
library(ggplot2)

ggplot(pca$x)+
  aes(PC1,PC2)+
  geom_point(col = my_cols)

```


The second major result figure is called a "loadings plot" of "variable contributions plot" or "weight plot"

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = rownames(pca$rotation)) +
  geom_col(fill = "steelblue") 

```

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```








